---
title: "GDP change analysis"
author: "Sophia Wu"
date: '`r Sys.time()`'
output:
  html_document:
    toc: yes
    toc_depth: 6
    toc_float: yes
  word_document:
    toc: yes
    toc_depth: '6'
  github_document:
    toc: yes
    toc_depth: 6
  pdf_document:
    toc: yes
    toc_depth: '6'
always_allow_html: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# *******Libraries*******
```{r message=FALSE, warning=FALSE}
library(tswge)
library(tidyverse)
library(ggplot2)
library(tseries)
library(kableExtra)
library(knitr)
library(plotly)
library(vars)
library(vars)
library(nnfor)
library(xts)
library(GGally)
library(astsa)
library(nnfor)
library(forecast)
```

#  *******Introduction*******
From Federal Reserve Economic Data, we see GDP changes every year.What is the economic cycle in the US? Are we able to predict the GDP change? Are there any significant variables causing GDP changes? Using Time series will help us analyze the important insights and predict the future trend.


#  *******Data Source*******
All data was downloaded from [Federal Reserve Economic Data (FRED) Repository](https://fred.stlouisfed.org/) It contains 75 years quarterly data from 1947 to 2021, 6 variables listed below:
6 columns as below:
*Date;
*GDP per capita;
*Income receipt;
*gross income;
*Profit before tax;
*gdp change.
Response variable is gdp change.

```{r}
# import US GDP data set
df <- read.csv("~/DS 6373 Time Series/TS Project2021/data/usgdp.csv")
df %>%  glimpse()
```


```{r}
# remove empty data, then plot dgp.change column. We may some outliers in the year 2020, which was during the COVID period. 
df <- na.omit(df)
plot(df$gdp.change)

```
```{r}
# We decided to get rid of the outlier which was affected by the COVID in the year 2020. Therefore, we deleted the data from the year 2020.
data <-head(df,-4)
plot(data$gdp.change)
```


```{r}
# matrix of scatter plots to check all of variables' correlation. It doesn't show much correlation between gdp change and other 4 variables.However, the top two important variables are GDP.PER.CAPITA and Income.receipt with correlation of -0.2 and -0.194
head(data)
ggpairs(data[2:6]) #matrix of scatter plots
```


#  ***check data stationarity ***
```{r, echo=FALSE}
#plot 75 years GDP Percent change from preceding period quarterly
plotts.sample.wge(data$gdp.change)

```


#  ***Condition 1: Constant Mean ***

* Wandering and oscillating behavior
* Only one realization not knowing much
* The assumption of constant mean seems like not appear to be violated.


#  ***Condition 2: Constant Variance ***
It is tough to confirm whether there is constant variance since we only have one realization. It seems like there is some tendency to make big jumps at the end of the period, indicating non-constant variance over time.


# ***Condition 3: Constant Autocorrelation ***
The first half of acf seems like little different from the second half of acf, but not much.
```{r}
# check the 1st half of acf for condition 3
acf(data$gdp.change[1:146])
```
```{r}
# check the 2nd half of acf for condition 3
acf(data$gdp.change[147:292])
```


#  ***Conclusion***
* Based on the above analysis, there is some evidence showing non-stationary.  But GDP change might be possibly stationary in the long term. We will analyze it using both stationary and non-stationary models.


#  ***Stationary model***

```{r}
# AIC selects ARMA(4,3) but BIC picks ARMA(0,2) showing ARMA(4,3) and ARMA(0,2) might be good fitting stationary models. 
aic5.wge(data$gdp.change, p=0:5, q=0:5)
```

```{r}
aic5.wge(data$gdp.change,type = "bic", p=0:5, q=0:5)
```
```{r}
aic5.wge(data$gdp.change,type = "aic", p=0:5, q=0:5)
```


# ***Model ID ARMA(4,3)***


```{r}
#aic pick ARMA(4,3), I'll test Stationary model ARMA(4,3)

ARMA43 = est.arma.wge(data$gdp.change,p=4,q=3)
```
```{r}
ljung.wge(ARMA43$res)
ljung.wge(ARMA43$res, K = 48)
f_ARMA43 = fore.arma.wge(data$gdp.change,phi = ARMA43$phi,n.ahead = 12,limits = F, lastn = T)
f_ARMA43short = fore.arma.wge(data$gdp.change,phi = ARMA43$phi,n.ahead = 2,limits = F, lastn = T)

```


```{r}
ASE_ARMA43 = mean((data$gdp.change[(length(data$gdp.change)-11):length(data$gdp.change)] - f_ARMA43$f)^2)

ASE_ARMA43

ASE_ARMA43short = mean((data$gdp.change[(length(data$gdp.change)-1):length(data$gdp.change)] - f_ARMA43short$f)^2)

ASE_ARMA43short
```


```{r}


#Compare Spectral Densities
sims = 5
SpecDen = parzen.wge(data$gdp.change, plot = "FALSE")
plot(SpecDen$freq,SpecDen$pzgram, type = "l", lwd = 6)

for( i in 1: sims)
{
   SpecDen2 = parzen.wge(gen.arma.wge(297, phi = ARMA43$phi, plot ="FALSE"), plot = "FALSE")
   lines(SpecDen2$freq,SpecDen2$pzgram, lwd = 2, col = "red")
}

```

```{r}

#Compare ACFs
sims = 5
ACF = acf(data$gdp.change, plot = "FALSE")
plot(ACF$lag ,ACF$acf , type = "l", lwd = 6)

for( i in 1: sims)
{
   ACF2 = acf(gen.arma.wge(297, phi = ARMA43$phi, plot = "FALSE"), plot = "FALSE")
   lines(ACF2$lag ,ACF2$acf, lwd = 2, col = "red")
}
```


```{r}

#Compare Generated Realizations 


ARMA43gen = gen.arma.wge(297,phi = ARMA43$phi, vara = ARMA43$avar)

plotts.sample.wge(ARMA43gen)
```


#  ***Non-stationary model***
```{r}
x=data$gdp.change
#difference the data
xd1.dif=artrans.wge(x,phi.tr=1)
#xd1.dif is the differenced data
plotts.sample.wge(xd1.dif)
```


```{r}
#run aic5.wge with difference data, it selects ARMA(5,1)
aic5.wge(xd1.dif,type = "bic")
```
```{r}
aic5.wge(xd1.dif,type = "aic")
```
```{r}
# AIC and BIC picks p=5, q=1
fitarima=arima(data$gdp.change,order=c(5,1,1),xreg=cbind(data$GDP.per.capita,data$Income.receipt,data$gross.income))
fitarima
AIC(fitarima) #1718.626
```

# ***Model ID ARIMA(5,1,1)***

```{r}

# we get ARIMA(5,1,1)
ARIMA51= est.arma.wge(xd1.dif,p = 5, q=1)

```
```{r}
ljung.wge(ARIMA51$res)
ljung.wge(ARIMA51$res, K = 48)
f_arima511=fore.arma.wge(data$gdp.change,phi=ARIMA51$phi,n.ahead = 12,limits = F, lastn = T)
f_arima511short=fore.arma.wge(data$gdp.change,phi=ARIMA51$phi,n.ahead = 2,limits = F, lastn = T)


ASElong = mean((data$gdp.change[(length(data$gdp.change)-11):length(data$gdp.change)] - f_arima511$f)^2)

ASElong

ASEshort = mean((data$gdp.change[(length(data$gdp.change)-1):length(data$gdp.change)] - f_arima511short$f)^2)

ASEshort
```


```{r}
ASE = mean((data$gdp.change[(length(data$gdp.change)-11):length(data$gdp.change)] - f_arima511$f)^2)

ASE
```
```{r}
#Compare Spectral Densities
sims = 5
SpecDen = parzen.wge(data$gdp.change, plot = "FALSE")
plot(SpecDen$freq,SpecDen$pzgram, type = "l", lwd = 6)

for( i in 1: sims)
{
   SpecDen3= parzen.wge(gen.arma.wge(295, phi = ARIMA51$phi, plot ="FALSE"), plot = "FALSE")
   lines(SpecDen3$freq,SpecDen3$pzgram, lwd = 2, col = "red")
}
```

```{r}
#Compare ARIMA51 ACFs
sims = 5
ACF = acf(data$gdp.change, plot = "FALSE")
plot(ACF$lag ,ACF$acf , type = "l", lwd = 6)

for( i in 1: sims)
{
   ACF2 = acf(gen.aruma.wge(319, phi = ARIMA51$phi, plot ="FALSE"), plot = "FALSE")
   lines(ACF2$lag ,ACF2$acf, lwd = 2, col = "red")
}
```


```{r}
#Compare Generated Realizations 
ARIMA51gen = gen.aruma.wge(297,phi = ARIMA51$phi, vara = ARIMA51$avar)
plotts.sample.wge(ARIMA51gen)
```

# ***Model ID ARUMA(5,1,0) S=27***
```{r}
# overfit for seasonality, we get system frequency 0.02 associated with period 27. 
plotts.sample.wge(x)
est.ar.wge(data$gdp.change,p = 12) 
factor.wge(data$gdp.change)
```


```{r}

#Transform  data with “seasonal difference”, which appears to be stationary, we'll find a stationary model for this realization
x=data$gdp.change
# difference the data
d1=artrans.wge(x,phi.tr=1)
#we try seasonal model with s=27
d1.27=artrans.wge(d1,phi.tr = c(rep(0,28),1))
                
```


```{r}
#Use tswge to model the transformed data, d1.27
aic5.wge(d1.27,p=0:5,q=0:5)
# aic picks an ARMA(5,0) as the first choice
```


```{r}
#In order to see if a lower order model could satisfactorily model the data, we use BIC
aic5.wge(d1.27,p=0:5,q=0:5,type='bic')
#BIC picks ARMA(5,0) as the second choice,We decide to use the ARMA(5,0) model chosen by AIC and BIC
```
```{r}
# AIC and BIC picks p=1,q=5
fitaruma=arima(data$gdp.change,order=c(5,1,0),xreg=cbind(data$GDP.per.capita,data$Income.receipt,data$gross.income))
fitaruma
AIC(fitarima) 
```

```{r}
#We estimate the parameters of the ARMA(5,0) model using est.ar.wge
x_27_arma=est.ar.wge(d1.27, p=5)
```



```{r}
ljung.wge(x_27_arma$res)
ljung.wge(x_27_arma$res, K = 48)
f_x27arma = fore.aruma.wge(data$gdp.change,s = 27,phi =x_27_arma$phi,n.ahead =12,limits = F, lastn = T)
f_x27armashort = fore.aruma.wge(data$gdp.change,s = 27,phi =x_27_arma$phi,n.ahead =2,limits = F, lastn = T)

ASE_x27arma= mean((data$gdp.change[(length(x)-11):length(x)] - f_x27arma$f)^2)

ASE_x27arma

ASE_x27arma_short= mean((data$gdp.change[(length(x)-1):length(x)] - f_x27armashort$f)^2)

ASE_x27arma_short
```
```{r}
#Compare Spectral Densities
sims = 5
SpecDen = parzen.wge(data$gdp.change, plot = "FALSE")
plot(SpecDen$freq,SpecDen$pzgram, type = "l", lwd = 6)

for( i in 1: sims)
{
   SpecDen4 = parzen.wge(gen.aruma.wge(297, phi = x_27_arma$phi,s=27, plot ="FALSE"), plot = "FALSE")
   lines(SpecDen4$freq,SpecDen4$pzgram, lwd = 2, col = "red")
}
```


```{r}
#Compare ACFs
sims = 5
ACF = acf(data$gdp.change, plot = "FALSE")
plot(ACF$lag ,ACF$acf , type = "l", lwd = 6)

for( i in 1: sims)
{
   ACF5 = acf(gen.aruma.wge(297, phi = x_27_arma$phi,s=27, plot = "FALSE"), plot = "FALSE")
   lines(ACF5$lag ,ACF5$acf, lwd = 2, col = "red")
}
```

# ***Multivariate analysis: VAR Model***

```{r}
data2 = data[1:292,]
gdpVar1 =VAR(cbind(data2$gdp.change,data2$GDP.per.capita,data2$Income.receipt), type="both", lag.max = 10)
AIC(gdpVar1)  #7177.838
preds_short = predict(gdpVar1,n.ahead = 2)
ASE_var1_st = mean((data$gdp.change[291:292] - preds_short$fcst$y1[,1])^2)
ASE_var1_st  #ASE for VAR model is 9.952455
plot(seq(1,292,1), data$gdp.change[1:292], type = "l",xlim = c(0,292), ylab = "gdp change", main = "Short_term gdp change Forecast")
lines(seq(291,292,1), preds_short$fcst$y1[,1], type = "l", col = "red")
```

```{r}
# VAR Model long-term forecast last 24 with much less ASE and AIC than short-term
data2 = data[1:268,]
gdpVar1 = VAR(cbind(data2$gdp.change,data2$GDP.per.capita,data2$Income.receipt), type="both", lag.max = 10)
AIC(gdpVar1)  
preds = predict(gdpVar1,n.ahead = 24)
ASEvar_lt=mean((data$gdp.change[269:292] - preds$fcst$y1[,1])^2)
ASEvar_lt 
plot(seq(1,292,1), data$gdp.change[1:292], type = "l",xlim = c(0,292), ylab = "gdp change", main = "Long_term gdp change Forecast")
lines(seq(269,292,1), preds$fcst$y1[,1], type = "l", col = "red")

```


# ***Multivariate analysis:Neural network***


```{r}

# MLP fit with 5 hidden nodes and 20 repetitions.
gdp.change=ts(data2$gdp.change)
GDP.per.capita=ts(data2$GDP.per.capita)
Income.receipt=ts(data2$Income.receipt)
xVar=data.frame(GDP.per.capita,Income.receipt)
set.seed(2)
fit.mlp=mlp(gdp.change,xreg=xVar)
fit.mlp
```





```{r}
## MLP fit with 5 hidden nodes and 20 repetitions, forecast combined using the median operator, MSE=7.08
xVar1=data.frame(GDP.per.capita=ts(data$GDP.per.capita),Income.receipt=ts(data$Income.receipt))
plot(fit.mlp)
## short-term forecast
fore.mlp_short = forecast(fit.mlp, h=2,  xreg = xVar1)
plot(fore.mlp_short)
ASE_nn_st = mean((data$gdp.change[291:292] - fore.mlp_short$mean)^2)
ASE_nn_st  
#Plot short_term forecast
plot(seq(1,292,1), data$gdp.change, type = "l",xlim = c(0,295), ylab = "gdp change", main = "Short_term GDP.Change Forecast from MLP")
lines(seq(291,292,1), fore.mlp_short$mean, type = "l", col = "red")
```
```{r}
## long-term forecast
fore.mlp = forecast(fit.mlp, h=24,  xreg = xVar1)
plot(fore.mlp)
ASE_nn_lt= mean((data$gdp.change[269:292] - fore.mlp$mean)^2)
ASE_nn_lt  
#Plot long_term forecast
plot(seq(1,292,1), data$gdp.change, type = "l",xlim = c(0,295), ylab = "gdp change", main = "Long_term  GDP.Change Forecast from MLP")
lines(seq(269,292,1), fore.mlp$mean, type = "l", col = "red")
```



# ***Ensemble model ***

```{r}
# long term ensemble 
ensemble  = (preds$fcst$y1[,1] + fore.mlp$mean)/2

#Plot
plot(seq(1,292,1), data$gdp.change, type = "l",xlim = c(0,292),  ylab = "gdp change", main = "Ensemble model long-term Forecast")
lines(seq(269,292,1), y=ensemble, type = "l", col = "RED")
ASE_ensemble = mean((data$gdp.change[269:292] - ensemble)^2)
ASE_ensemble  
```
```{r}
## short term ensemble model
ensemble  = (preds_short$fcst$y1[,1] + fore.mlp_short$mean)/2

#Plot
plot(seq(1,292,1), data$gdp.change, type = "l",xlim = c(0,292),  ylab = "gdp change", main = "Ensemble model GDP.Change Forecast")
lines(seq(291,292,1), y=ensemble, type = "l", col = "RED")
ASE_ensemble = mean((data$gdp.change[291:292] - ensemble)^2)
ASE_ensemble  #4.8

```

# ***Model Comparison***

```{r}
#For the short-term forecast, we may see ARIMA(5,1,0)  with s=27 shows better performance with lowest ASE 14.18755;
#For the long-term forecast, the ensemble model shows the better performance with the lowest ASE 4.78.

print(' *ARMA(4,3)      Short-term ASE 37.968,  Long-term ASE 6.879'  )
print(' *ARIMA(5,1)     Short-term ASE 35.09,   Long-term ASE 6.515'  )
print(' *ARUMA(5,1,0)   Short-term ASE 14.187,  Long-term ASE 6.365'  )
print(' *VAR            Short-term ASE 25.895,  Long-term ASE 5.15'  )
print(' *MLP            Short-term ASE 15.16,   Long-term ASE 6.27'  )
print(' *Ensemble       Short-term ASE 16.259,  Long-term ASE 4.78'  )
```

# ***Summary***

* GDP data is very difficult to predict, and models are not able to capture variance in this data.
* For the short-term forecast, we may see ARIMA(5,1,0)  with s=27 shows better performance with lowest ASE 14.18755;
* For the long-term forecast, the ensemble model shows the better performance with the lowest ASE 4.78.
* After all, we can't simply say which model is better than the other as they perform differently in the different periods.


